Methods for Analyzing Models of Rod-Shaped Bacteria

### 2025-07-14 paper/main.tex
restructure; try to unify paragraphs and subsections
add and rename table of contnents
more restructuring

\tableofcontents

@@

%###################################################################################################
\section{Fundamental Assumptions on the Spatial Representation of Rod-Shaped Bacteria}
% TODO CITATIONS
This work makes use of a few but powerfull assumptions concerning the mathematical representation of
the cells in question.
In order to describe the spatial dynamics of individual agents, we need to define their position,
shape and how they interact with each other.
In order to provide as much flexibility as possible when constructing new models, we restrict
ourselves to assumptions concerning the position such that researchers can freely experiment with
various shapes and interaction types.
Furthermore, it is very important to strike a good balance between the restrictiveness of our
assumptions and generality in order to provide meaningful results and still be able to apply our
approach to a broader class of models.
Taking into account the natural shape of rod-like bacteria, it is clear that they can not be
represented with a singular point $\vec{v}$ in space.
We could extend this positional definition with a vector $\vec{w}$, indicating the direction of
extension but this would restrict ourselves to fully straight rod-shaped bacteria.
However, multiple experiments have shown that MReB is responsible for the growth and maintenance of
the elongated shape of bacteria~\cite{Erickson2001}.
By means of this protein, rod-shaped bacteria can bend elastically on short time-scales and
deform plastically over longer periods of time.
Our goal is to provide a model which can encapsulate these effects.
Thus, we extend our model from a single point to multiple vertices which are part of a 1-dimensional
polygon with two endpoints $\vec{v}_i$.
In essence, we discretize the (possibly non-straight) rod of the agent.
Figure~\ref{fig:mechanics-bacterium} shows a visualization of this approach.
The angles $\alpha_i$ between vertices $\vec{v}_{i-1},\vec{v}_i$ can be used when modeling the
dynamics of these bacteria.

@@

%###################################################################################################
\section{Data Extraction}
\label{section:fitting-and-comparing-masks}

@@

% --------------------------------------------------------------------------------------------------
\paragraph{Benchmarking the Extraction Algorithm}
\label{subsection:benchmarking-extraction-algorithm}

@@

%###################################################################################################
\section{Visualization}
\label{section:generating-masks-and-images}
\begin{itemize}
    \item use pyvista (VTK) for Visualization
    \item always do 3D render. This way, we can have overalp between agents and produce more
        realistic masks.
    \item Assign unique color values for idividual agents.
        We have a 1:1 mapping between agent identifier and color! (for each simulation different
        though)
    \item use parallel projection; this should be most close to what microscopes measure due to the
        relatively large distance between object and lens.
    \item use some noise to smear out the rods; this is already similar to microscopic images; use
        more like this and deal with real-world defects to make it more realistic (future, not now)
    \item Figure~\ref{fig:progression-image-generation} shows steps in generating image.
    \item Reflectivity, placement of light etc. can already introduce some heterogeneity in the
        visual of the rod.
    \item Plan to implement noise on the mesh itself; make it have rough surface (like real rods as
        seen under electron microscopy)
\end{itemize}

@@

%###################################################################################################
\section{Parameter Estimation}
In order to effectively estimate the parameters of our model, we need to set up a cost function
which can be minimized.
In doing so, we hope to identify all parameters, given relevant experimental data.
We use microscopic images and construct a novel algorithm that can extract the position of the agent
from a provided cell-mask.
These cell-masks are easily obtainable by either using existing segmentation tools or manually
annottating images.
Following this, we can define a cost-function and an optimization scheme to minimize said function.

% --------------------------------------------------------------------------------------------------
\subsection{Constructing a Loss Function}

@@

\paragraph{Comparison with Semantic Segmentation}
\label{paragraph:comparison-semantic-segmentation}
\cite{Jadon2020}

\paragraph{Comparison across Generations}
\label{paragraph:comparison-across-generations}

@@

\section{Applications}
% --------------------------------------------------------------------------------------------------
\subsection{A Mechanical Model of Rod-Shaped Bacteria}
\label{subsubsection:mechanical-model-rod-shaped-bacteria}

@@

% --------------------------------------------------------------------------------------------------
\subsection{Interactions}
\label{subsection:mechanical-model-interactions}

@@

% --------------------------------------------------------------------------------------------------
\subsection{Cycle}
\label{subsection:mechanical-model-cycle}

>>>>

\pagebreak
\renewcommand{\contentsname}{Temporary! Table of Contents}
\tableofcontents
\pagebreak

@@

%###################################################################################################
\section{Method Description}
\subsection{Fundamental Assumptions on the Spatial Representation of Rod-Shaped Bacteria}
% TODO CITATIONS
This work makes use of a few but powerfull assumptions concerning the mathematical representation of
the cells in question.
In order to describe the spatial dynamics of individual agents, we need to define their position,
shape and how they interact with each other.
In order to provide as much flexibility as possible when constructing new models, we restrict
ourselves to assumptions concerning the position such that researchers can freely experiment with
various shapes and interaction types.
Furthermore, it is very important to strike a good balance between the restrictiveness of our
assumptions and generality in order to provide meaningful results and still be able to apply our
approach to a broader class of models.
Taking into account the natural shape of rod-like bacteria, it is clear that they can not be
represented with a singular point $\vec{v}$ in space.
We could extend this positional definition with a vector $\vec{w}$, indicating the direction of
extension but this would restrict ourselves to fully straight rod-shaped bacteria.
However, multiple experiments have shown that MReB is responsible for the growth and maintenance of
the elongated shape of bacteria~\cite{Erickson2001}.
By means of this protein, rod-shaped bacteria can bend elastically on short time-scales and
deform plastically over longer periods of time.
Our goal is to provide a model which can encapsulate these effects.
Thus, we extend our model from a single point to multiple vertices which are part of a 1-dimensional
polygon with two endpoints $\vec{v}_i$.
In essence, we discretize the (possibly non-straight) rod of the agent.
Figure~\ref{fig:mechanics-bacterium} shows a visualization of this approach.
The angles $\alpha_i$ between vertices $\vec{v}_{i-1},\vec{v}_i$ can be used when modeling the
dynamics of these bacteria.

@@

%###################################################################################################
\subsection{Data Extraction}
\label{section:fitting-and-comparing-masks}

@@

% --------------------------------------------------------------------------------------------------
\paragraph{Benchmarking the Extraction Algorithm}
\label{paragraph-extraction-algorithm}

@@

%###################################################################################################
\subsection{Visualization}
\label{section:generating-masks-and-images}
\begin{itemize}
    \item use pyvista (VTK) for Visualization
    \item always do 3D render. This way, we can have overalp between agents and produce more
        realistic masks.
    \item Assign unique color values for idividual agents.
        We have a 1:1 mapping between agent identifier and color! (for each simulation different
        though)
    \item use parallel projection; this should be most close to what microscopes measure due to the
        relatively large distance between object and lens.
    \item use some noise to smear out the rods; this is already similar to microscopic images; use
        more like this and deal with real-world defects to make it more realistic (future, not now)
    \item Figure~\ref{fig:progression-image-generation} shows steps in generating image.
    \item Reflectivity, placement of light etc. can already introduce some heterogeneity in the
        visual of the rod.
    \item Plan to implement noise on the mesh itself; make it have rough surface (like real rods as
        seen under electron microscopy)
\end{itemize}

@@

%###################################################################################################
\subsection{Parameter Estimation}
In order to effectively estimate the parameters of our model, we need to set up a cost function
which can be minimized.
In doing so, we hope to identify all parameters, given relevant experimental data.
We use microscopic images and construct a novel algorithm that can extract the position of the agent
from a provided cell-mask.
These cell-masks are easily obtainable by either using existing segmentation tools or manually
annottating images.
Following this, we can define a cost-function and an optimization scheme to minimize said function.

% --------------------------------------------------------------------------------------------------
\subsubsection{Challenges of Constructing a Loss Function}

@@

% --------------------------------------------------------------------------------------------------
\subsubsection{Relationship to Semantic Segmentation}
\label{paragraph:comparison-semantic-segmentation}
\cite{Jadon2020}

% --------------------------------------------------------------------------------------------------
\subsubsection{Comparison across Generations}
\label{paragraph:comparison-across-generations}

@@

% --------------------------------------------------------------------------------------------------
\section{Application to Mechanics of Rod-Shaped Bacteria}
\label{subsubsection:mechanical-model-rod-shaped-bacteria}

@@

% --------------------------------------------------------------------------------------------------
\subsection{Physical Interactions}
\label{subsection:mechanical-model-interactions}

@@

% --------------------------------------------------------------------------------------------------
\subsection{Cell-Cycle}
\label{subsection:mechanical-model-cycle}
